---
title: "mtcars regression and chi-square analyses with plots"
author: "Prince Parker"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
---

#Setup chunk
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

###Code structure
1. Load packages and view dataset
2. Run linear regression model of weight and mpg
3. Plot regression model
4. Run Chi-Square Analysis of Transmission Type and Engine



#Load necessary packages
```{r}
#Load packages#
library(tidyverse)
library(ggplot2)
library(readr)
library(knitr)

```

#Read in mt cars csv and display the first 5 lines#
```{r}
#Read in mt cars csv

data("mtcars")

#Display the first 5 lines
head(mtcars)

```
The scatterplot below shows the relationship between **weight (wt) and miles per gallon (mpg)** for the cars in the `mtcars` dataset. The fitted **linear regression model** suggests that for every additional **1000 lbs** of weight, **MPG decreases by approximately `r round(coef(lm(mpg ~ wt, data = mtcars))[2], 2)`**.

The regression equation is:

> **MPG = `r round(coef(lm(mpg ~ wt, data = mtcars))[1], 2)` + `r round(coef(lm(mpg ~ wt, data = mtcars))[2], 2)` * Weight`**

The R-squared value, indicating how well weight explains MPG variation, is **`r round(summary(lm(mpg ~ wt, data = mtcars))$r.squared, 3)`**.

## **Regression Analysis: MPG vs. Weight**
```{r}
# Run simple linear regression
model <- lm(mpg ~ wt, data = mtcars)

# Print summary of the regression model
summary(model)


```


```{r}
#Plot mtcars scatterplot with regression line
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(color = "red", shape = "triangle", size = 3) +  # Scatter plot
  geom_smooth(method = "lm", color = "green", se = TRUE) +  # Regression line with confidence interval
  labs(title = "Regression of MPG on Weight", x = "Weight (1000 lbs)", y = "Miles per Gallon") +
  theme_minimal()


```

We conducted a **Chi-Square test** to examine the relationship between **Transmission** and **Engine type**. The results indicate a **Chi-Square statistic of `r round(chisq_test$statistic, 2)`**, with a **p-value of `r signif(chisq_test$p.value, 3)`**.

Based on a significance level of **0.05**, we **`r ifelse(chisq_test$p.value < 0.05, "reject", "fail to reject")`** the null hypothesis. This suggests that there **`r ifelse(chisq_test$p.value < 0.05, "is", "is no")`** significant association between transmission type and engine type.
###Conduct Chi-Squared Test
```{r}

# Create a contingency table
table_mtcars <- table(mtcars.viz$Transmission, mtcars.viz$Engine)

# Perform the Chi-Square test
chisq_test <- chisq.test(table_mtcars)

# View results
print(chisq_test)

```
```{r}
summary(table_mtcars)
```

<!-- No association between Transmission and Engine type (p>.05) -->
###Results
- Weight predicts car miles per gallon
- There was no association between transmission type and engine


When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

